\section{Introduction}
\label{sec:introduction}

\subsection{From bounded memory to information frictions: how the project was scoped}
\label{sec:intro-scoping}

The original plan for this thesis (\citealp[Section~2]{calvano2020} as the baseline; bounded-memory and frictional extensions on top) was to relax two distinct idealisations of the Calvano environment: the assumption that agents have unbounded memory of past prices, and the assumption that observations of those prices are instantaneous, noiseless and synchronous. Of these, the first turns out to be computationally infeasible at the compute budget of a bachelor thesis. Quantifying \emph{why} is informative for what follows; this subsection states the argument explicitly.

Each agent's state at time $t$ is the joint price vector over the previous $k$ periods, so the state space has cardinality $\lvert \mathcal{S}\rvert = m^{nk}$, where $n$ is the number of firms and $m$ the number of grid prices. The Calvano baseline uses $(n, k, m) = (2, 1, 15)$, giving $\lvert \mathcal{S}\rvert = 225$ and a per-agent Q-table of $m^{nk+1} = m \cdot \lvert \mathcal{S}\rvert = 3{,}375$ cells --- a manageable problem.

The cost of an additional unit of memory is non-trivial. With $n=2$ and $m=15$, moving from $k=1$ to $k=2$ inflates the state space to $m^{nk}=15^{4} = 50{,}625$, a $225$-fold increase, and the per-agent Q-table to $759{,}375$ cells. The memory footprint is still manageable. The exploration coverage is not. \citet[footnote to eq.~21]{calvano2020} give the expected number of visits to a sub-optimal Q-cell, under $\varepsilon$-greedy exploration, as
\begin{equation}
    \nu \;=\; \frac{(m-1)^{n}}{m^{kn(n+1)} \cdot \bigl(1 - e^{-\beta(n+1)}\bigr)}.
    \label{eq:nu-visit-count}
\end{equation}
Holding $\beta$ and the per-session iteration budget fixed and incrementing $k$ by one with $n=2$ multiplies the denominator of \eqref{eq:nu-visit-count} by $m^{n(n+1)} = m^{6}$. At $m=15$ that is $15^{6} \approx 1.14 \times 10^{7}$. Each Q-cell receives, on average, ten million times fewer learning updates --- so either the iteration budget would have to grow by a factor of $10^{7}$ (pushing session runtimes from minutes to weeks), or $\beta$ would have to grow correspondingly (which destroys the late-session low-$\varepsilon$ exploitation regime needed for collusive behaviour to crystallise). Neither is feasible inside this project.\todo{The math here is the standard Calvano argument; double-check that the exponent on $m$ in the denominator reads $kn(n+1)$ rather than the slightly different forms quoted in some surveys.}

The implication is a re-scoping. The frictions implemented in this thesis --- observation latency, noisy monitoring and asynchronous actions --- all preserve the per-agent state cardinality at $m^{2}=225$.\footnote{Latency and noise change \emph{which} pair of prices is in the perceived state, not how many possible pairs there are. Asynchrony does not change the state space at all.} They are therefore tractable at the same compute scale that runs the Calvano baseline. The cost of this re-scoping is that the question ``does longer memory protect collusion?'' cannot be answered here, only ``does relaxing the information and timing assumptions break it?'' This is precisely the question the proposal listed as central.

\subsection{Why algorithmic collusion matters}
\label{sec:intro-motivation}

Reinforcement-learning algorithms increasingly set prices in real-world online markets, and at least one ex-post empirical study finds material price effects of algorithmic adoption: \citet{assad2024} document non-trivial markup increases in the German retail gasoline market in the period when major chains adopted algorithmic pricing software. Whether this reflects active anti-competitive behaviour, or simply more efficient unilateral pricing, is the central interpretive question for competition policy.

\citet{calvano2020} provide the foundational simulation evidence on the first interpretation. Two independent tabular Q-learning agents in a repeated Bertrand duopoly with logit demand reliably converge on prices substantially above the static-Bertrand Nash equilibrium. The agents discover, without communication, reward--punishment strategies that sustain joint high prices: deviations are detected and met with temporary price wars, after which prices return to the supracompetitive level. The convergence is robust to a wide range of hyperparameter choices. \citet{oecd2017} review the policy implications.

The Calvano result depends on the same idealisations that have always made formal analysis of repeated-game collusion tractable but conceptually fragile: instantaneous, noise-free, simultaneous observation of the rival's price. Real markets fail each assumption. Price changes propagate through systems and consumer-facing channels with a lag; observed prices are noisy because of catalogue errors, channel-specific variation, and mid-period promotions; and pricing decisions are typically asynchronous, with different firms reviewing prices on different cadences. Each feature has classical analytic precedent in the industrial-organisation literature on tacit collusion. \citet{green1984} formalise imperfect monitoring and show that even tacit collusion under noisy demand signals can sustain supracompetitive prices, but only at the cost of recurrent equilibrium-path price wars triggered by adverse signals. \citet{maskin1988} formalise alternating-move pricing and characterise its Markov-perfect equilibria, finding both Edgeworth-cycle and focal-price equilibria. Whether learned (rather than equilibrium) collusion behaves the same way under the same frictions is the empirical question.

\subsection{Existing work and the gap}
\label{sec:intro-litreview}

A growing literature studies frictional extensions of the Calvano baseline, but each in isolation and under heterogeneous baselines. \citet{calvano2021} introduce imperfect monitoring through stochastic demand in a Cournot setting, finding that algorithmic collusion survives moderate monitoring noise but is disrupted at higher levels. \citet{klein2021} treats sequential price-setting as the base game rather than as a deviation from the simultaneous benchmark, showing that Q-learners in alternating-move pricing also converge on supracompetitive prices, sometimes via Edgeworth-cycle dynamics; the same paper appears in earlier form as a Tinbergen Institute working paper.\footnote{\citet{klein2021} corresponds to the working-paper version Klein (2018), Tinbergen Institute Discussion Paper TI 2018-056/VII.} \citet{calvano2023} ask whether the high prices that Q-learners converge on are ``genuine'' (sustained by punishment) or ``spurious'' (an artefact of the action-selection rule), and propose off-path deviation tests to distinguish the two. \citet{asker2024} vary the learning protocol itself --- algorithm choice, exploration rule, learning rate --- rather than the information environment. \citet{zhang2025} considers Laplacian observation noise as a regulatory instrument and argues that injecting bounded noise into the rival-price signal can break learned collusion in their environment.

What none of these papers does is to introduce two or more frictions \emph{simultaneously} in a single controlled environment and ask whether they interact. Real markets do not feature one friction at a time. A regulator considering noise injection as a policy tool, for example, would want to know whether the policy's effectiveness depends on whatever observation latency or pricing-cadence asynchrony the market already has. The present paper attempts to fill this gap.

\subsection{Research questions}
\label{sec:intro-rqs}

Within the otherwise unchanged \citet{calvano2020} environment ($n=2$, $m=15$, $k=1$, logit demand, baseline $\alpha$, $\beta$, $\delta$), this thesis addresses:
\begin{enumerate}[label=RQ\arabic*., leftmargin=*]
    \item How does \emph{observation latency} (delayed observation of the rival's last-period price) affect the converged level of collusion? Is there a threshold latency beyond which collusion collapses?
    \item How does \emph{noisy monitoring} (Gaussian misobservation of the rival's last-period price) affect collusion? Does it produce gradual decay, or a Green--Porter-style equilibrium with recurrent price wars, or eventual Nash?
    \item How does \emph{asynchronous price-setting} (Maskin--Tirole alternation between movers) affect collusion?
    \item Do these frictions \emph{interact} when imposed jointly, in particular non-linearly such that pairs or triples of moderate frictions break collusion that any one of them alone leaves intact?
\end{enumerate}

\subsection{Contributions}
\label{sec:intro-contributions}

This paper makes three contributions. First, it provides a like-for-like comparison of three frictions within a single controlled environment, with replication counts (250, 500 and 1000 sessions per cell) sufficient to give cross-session standard deviations below $0.30$ on $\Delta\in[0,1]$ and standard errors of the mean below $0.02$ on the headline cells. The single-friction sweeps reveal three qualitatively different shapes of the $\Delta(\cdot)$ curve --- a cliff (latency), a smooth monotone decline (noise), and a step-then-plateau (asynchrony) --- a finding worth documenting in its own right.

Second, the paper presents a $2\times 2\times 2$ factorial experiment, replicated at three session-counts, that crosses all three frictions at moderate non-saturating levels. Two interaction effects survive replication. Latency and noise compound super-multiplicatively: the joint effect is stronger than the product of single-friction ratios would predict. And asynchrony, despite reducing collusion in isolation, partially \emph{rescues} it when imposed on top of either or both of the other two frictions. The latter is an opposite-sign interaction: the same intervention has different signs as a stand-alone policy versus as an addition to an already-frictional environment.

Third, the paper draws a tentative policy implication. An information-degradation rule (mandated observation delay, mandated noise) and a structural-timing rule (mandated staggered pricing) appear to be policy \emph{substitutes}, not complements: imposing the structural-timing rule in a market that already has informational frictions can partially undo the effect of those frictions on learned collusion. Regulators considering combinations of these instruments should not assume the effects add.

\subsection{Structure}
\label{sec:intro-structure}

Section~\ref{sec:methodology} describes the baseline environment, the Q-learning rule, the three friction extensions, and the unified factorial implementation, including each design choice and any deviation from the proposal. Section~\ref{sec:results-singles} presents the single-friction sweeps. Section~\ref{sec:results-factorial} presents the factorial experiment. Section~\ref{sec:discussion} interprets the findings. Section~\ref{sec:limitations} states limitations and points to extensions.

\paragraph{Note on draft conventions.}
This is a working draft. Footnotes in coloured ``\texttt{[draft note]}'' format flag claims, references or numerical results that need verification or refinement before the document is finalised. They are not part of the intended final text.\todo{Remove this paragraph and convert all draft notes to either resolved-and-deleted or proper footnotes before submission.}
